### 2025-06-15 talk/cmo25.tex
cmo25 talk revised

\item Bonus question: How do we know if our model is statistically adequate, or needs more work? Hint: we need {\bf performance benchmarks}, i.e., comparison with non-mechanistic machine learning methods. If there is performance shortfall, we need diagnostics to figure out where and how.

    \end{enumerate}
\end{frame}

@@

\begin{frame}{Inference challenges in population dynamics}

@@

  1--5 are largely solved, from a methodological perspective.\\
  6 is our immediate topic.\\
  7 is exciting but not the focus of this talk.

@@

    \begin{itemize}
    \item  Iterated particle filtering via \texttt{mif2} in the R package \texttt{pomp} enables Masters-level statisticians to do plug-and-play likelihood-based inference for nonlinear, non-Gaussian, partially observed dynamic systems:

@@

  Asfaw et al ({\it J. Open Source Software}, 2024)
  \url{https://github.com/spatPomp-org}

  \bi
\item Permits specification of arbitrary non-linear non-Gaussian spatiotemporal POMP models.
\item Adds spatial structure to \texttt{pomp} \citep{king16}.
\item Monte Carlo algorithms designed to address the \myemph{curse of dimensionality}.
  \item Algorithms compile via C to assist performance.
\item Statistically efficient, likelihood-based inference despite methods being simulation-based.
  \ei

@@

\end{center}
\bi
\item 
  Reportably infectious individuals, $\mathrm{I}_u$ for city $u$, are included in the delayed reporting compartment, $\mathrm{C^a_u}$.
  \item 
An individual arriving at $\mathrm{C}_u$ is a case report for city $u$.
\item
  Individuals in $\mathrm{A}_u$ are not reportable and transmit at a reduced rate.
  \item 
Travel occurs to and from $\mathrm{T}$, based on 2018 data from Tencent.
%\item Many models have been proposed. We want a framework to implement and evaluate various scientific hypotheses.
\ei

@@

  \resizebox{12cm}{!}{
\begin{tabular}{c|rr|l}
 & loglik & df & description
\\
\hline
$M_1$ &
  $-\infty$ & 10 &
  SEAIR model, parameter values and mobility of Li et al (2020)
\\
$M_2$ &  -14240.5 & 10 &
  Adjusted mobility and measurement in $M_1$
\\
$M_3$ &
 -11257.9  & 374 &
  Independent identically distributed negative binomial
\\
$M_4$ &
 -10825.3 & 375 &
  Autoregressive negative binomial
\\
$M_5$ &
  -9088.2&
  12 &
  Adding overdispersed dynamics to $M_2$ and refitting
\\
$M_6$ &
  -9116.5&
  10 &
  Latent and infectious periods unchanged by lockdown in $M_5$
\\

@@

\begin{frame}{Data analysis progression}
  \bi
\item Some data errors meant that the original data was incompatible with the model.
\item Tracked down by assessing conditional log-likelihood for each unit and each time.
\item Even then, the likelihood was poor compared to non-mechanistic benchmarks. Including over-dispersion fixed that.
\item Some weak identifiability remained. extra assumptions can lead to plausible parameter estimates without much loss of model fit.
\item Mild outliers remain, e.g., Wuhan on day 18.  
  \ei

@@

  \begin{itemize}
\item BPF is the best available filter for this COVID model.
\item Similar results hold for other models (Ionides et al, {\it JASA}, 2023).
  \end{itemize}

@@

    \item Two recent case studies (Wheeler et al, 2023; Li et al, 2023) demonstrate data analysis using IBPF. Code and data are provided via R packages extending \code{spatPomp}.

@@

\nocite{asfaw24,bjornstad01,breto19,grenfell04,he10,katzfuss19,ionides21,ionides24,lee20,li20,ng02,ning23-ibpf,park20,rebeschini15,wheeler24,li24}

>>>>

\item Bonus question: How do we know if our model is statistically adequate, or needs more work?

  \begin{itemize}
  \item Hint: we need {\bf performance benchmarks}, i.e., comparison with non-mechanistic machine learning methods. If there is performance shortfall, we need diagnostics to figure out where and how.
  \end{itemize}
    \end{enumerate}
\end{frame}

@@

\begin{frame}{Inference challenges in population dynamics: A brief history}

@@

  1--5 are largely solved, from a methodological perspective.\\
  \myemph{6 is our immediate topic.}\\
  7 is exciting but not the focus of this talk.

@@

    \begin{itemize}
    \item  \myemph{Iterated particle filtering} via \texttt{mif2} in the R package \texttt{pomp} \citep{king16} enables plug-and-play likelihood-based inference for nonlinear, non-Gaussian, partially observed Markov process (POMP) models.

      \item Allows consideration of arbitrary POMP models for low-dimensional time series data.

      \item Accessible for Masters-level statisticians:

@@

  Asfaw et al (2024, {\it J. Open Source Software})
  \url{https://github.com/spatPomp-org}

  \bi
\item Permits specification of arbitrary non-linear non-Gaussian spatiotemporal POMP (SpatPOMP) models.
\item \texttt{spatPomp} spatial structure to \texttt{pomp}.
\item Includes Monte Carlo algorithms designed to address the \myemph{curse of dimensionality}.
  \item Algorithms compile via C to assist performance.
\item Statistically efficient, likelihood-based inference despite methods being simulation-based.
  \ei

@@

\end{center}
  \bi
  \item $U=373$ cities (spatial units) labeled $u=1,\dots,U$.
\item 
  Reportably infectious individuals, $\mathrm{I}_u$ for city $u$, are included in the delayed reporting compartment, $\mathrm{C^a_u}$.
  \item 
An individual arriving at $\mathrm{C}_u$ is a case report for city $u$.
\item
  Individuals in $\mathrm{A}_u$ are not reportable and transmit at a reduced rate.
  \item 
Travel occurs to and from $\mathrm{T}$, based on 2018 data from Tencent.
%\item Many models have been proposed. We want a framework to implement and evaluate various scientific hypotheses.
\ei

@@

  \resizebox{12cm}{!}{
\begin{tabular}{c|rr|l}
 & loglik & df & description
\\
\hline
$M_1$ &
  $-\infty$ & 10 &
  SEAIR model, parameter values and mobility of Li et al (2020)
\\
$M_2$ &  -14240.5 & 10 &
  Adjusted mobility and measurement in $M_1$
\\
$M_3$ &
 -11257.9  & 374 &
  Independent identically distributed negative binomial
\\
$M_4$ &
 -10825.3 & 375 &
  Autoregressive negative binomial
\\
$M_5$ &
  -9088.2&
  12 &
  Adding overdispersed dynamics to $M_2$
\\
$M_6$ &
  -9116.5&
  10 &
  Latent and infectious periods unchanged by lockdown in $M_5$
\\

@@

\begin{frame}{Data analysis progression}
  \bi
\item Some data errors meant that the original data was incompatible with the model.
\item Tracked down by assessing conditional log-likelihood for each unit and each time.
\item Even then, the likelihood was poor compared to non-mechanistic benchmarks. Including over-dispersion fixed that.
\item Some weak identifiability remained. extra assumptions can lead to plausible parameter estimates without much loss of model fit.
\item Less severe outliers remain, e.g., Wuhan on day 18.  
  \ei

@@

  \begin{itemize}
\item Block particle filter (BPF) beats alternatives for this COVID model.
\item Similar results hold for other models (Ionides et al, {\it JASA}, 2023).
  \end{itemize}

@@

    \item Two recent case studies (Wheeler et al, 2024; Li et al, 2024) demonstrate data analysis using IBPF. Code and data are provided via R packages extending \code{spatPomp}.

@@

%\nocite{asfaw24,bjornstad01,breto19,grenfell04,he10,katzfuss19,ionides23,ionides24,lee20,li20,ng02,ning23-ibpf,park20,rebeschini15,wheeler24,li24}

\nocite{asfaw24,bjornstad01,grenfell04,ionides23,ionides24,li20,ng02,ning23-ibpf,park20,rebeschini15,wheeler24,li24}
